---
title: "Results"
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide

library(tidyverse)
library(dplyr)
library(psych)
library(knitr)
library(psych)

omaha_bridges <- read.csv("omaha_bridges.csv")

fil_dat <- omaha_bridges %>%
  select(c(1,2,5,15,21,22,25, 28:29,31:32, 68:77,88,103,107,112,116, 122:125))
col <- data.frame(colnames(fil_dat))

fil_dat <- fil_dat %>%
  mutate(across(everything(), ~ ifelse(. == "N", NA, .)))

bridge_cond1 <- fil_dat %>%
 filter(
   BRIDGE_CONDITION == "P"
 ) %>%
   select(c(1,8,9,10,12:16,21,26,27))

high_traffic <- bridge_cond1 %>%
  filter (
    ADT_029 > 5000 & TRAFFIC_LANES_ON_028A > 2
  )

desc_brid_cond <- describe(bridge_cond1)
desc_h_t <- describe(high_traffic)
```

The first step to the formal analysis is to find the descriptive statistics from the bridges that are in poor condition.

library(psych)

```{r}
#| echo: false
#| message: false
#| warning: false
kable(desc_brid_cond[, c("mean", "sd", "n","median","se")], caption = "Bridges with Poor Rating Descriptive Statistics")


```

From the table above we can see that the average year of the bridges built in Omaha, NE is 1955. We can also see fro this data that all of our structural condition ratings are low as the average is below 4 which tells us that the structural condition of the all of these bridges is below par. We can also see that the average daily traffic(ADT_029) median is 2,027 cars per day. We will now look at the descriptive statistics for the high traffic data subset.

```{r}
#| echo: false
#| message: false
#| warning: false
kable(desc_h_t[, c("mean", "sd", "n","median","se")], caption = "High Traffic Bridges with a Rating of Poor")
```

These bridges are at risk of collapse and they also have the highest traffic flow among at risk bridges. These bridges are further at risk due to the fact that the mean percentage of truck traffic is 26% which is very high compared to most bridges in Omaha, NE. We can see this in the descriptive statistics as the mean of the structural conditions is below 3 which is lower than if the bridges were just at poor condition. We can also see that the average year built of the bridges is slightly newer at 1969. Finally, we can see that median of the average daily traffic is 34353 which is much higher than the previous subset. The next step is to look at the proportion of bridges in Omaha that have a poor condition and that have a poor condition under high traffic.

```{r}
#| echo: false
#| message: false
#| warning: false
percents <- data.frame(
  "bridges at risk" = nrow(bridge_cond1)/nrow(fil_dat),
  "High traffic bridges at risk" = nrow(high_traffic)/nrow(fil_dat))
kable(percents, col.names = c("Bridges at Risk", "High Traffic Bridges at Risk"))
```

From this table we can see that the percent chance that a bridge in Omaha, NE is in poor condition is 5%. Along with that, the percent chance that a bridge in poor condition has high traffic is 1.4%. This tells us that we have a minimal chance of coming across a bridge that is at risk of collapse by structural ratings. This was a solution primarily based on descriptive statistics, we will now create a statistical model to predict a collapse of a bridge. To do this task we must create a collapse variable. This will be done by creating a binary variable that depends on if the bridge condition is poor and is the structural evaluation of the bridge in less than 4. These two variables give us the best insight into if a bridge is at risk of collapse, hence they will be used to create the variable. The data that we have also contains many "NA" values, to take care of this issue will will impute the data, or estimate values for the "NA" values based on information in the data set. The new imputed data set is what we will use to form our model.

The model we will use will be the following:

$$
\log\left(\frac{P(\text{collapse_risk} = 1)}{1 - P(\text{collapse_risk} = 1)}\right) 
= \beta_0 + \beta_1 \cdot \text{DECK_COND_058} + \beta_2 \cdot \text{OPERATING_RATING_064} 
+ \beta_3 \cdot \text{YEAR_BUILT_027}
$$

The model above is a generalized linear model, as our response variable is non - normal ( Binomial), this is done using the logit link function.

```{r}
#| echo: false
#| message: false
#| warning: false

Omaha <- fil_dat %>%
  mutate(collapse_risk = ifelse(BRIDGE_CONDITION == "P" & STRUCTURAL_EVAL_067 <= 4, 1, 0)) %>%
  select(c(1,8,12,18,26,31))

library(VIM)
one <- kNN(Omaha, k = 5)


library(caret)
train_indices <- createDataPartition(one$OBJECTID, 
                                     p = 0.8, 
                                     list = FALSE)

train_data <- one[train_indices, ]

# Testing data
test_data <- one[-train_indices, ]


# Logistic regression model
model1 <- glm(collapse_risk ~as.numeric(DECK_COND_058) + OPERATING_RATING_064 + as.numeric(YEAR_BUILT_027), 
             family = binomial(link = "logit"), data = train_data)

library(gt)
library(broom)

tidy(model1) %>%
  gt() %>%
  tab_header(
    title = "Logistic Regression Summary"
  )

```

From this model we can see that all variables are significant and should produce accurate results in predicting our response variable on testing data. We will now predict the response on a test data set using this model.

```{r}
#| echo: false
#| message: false
#| warning: false
predicted_probabilities <- predict(model1, newdata = test_data, type = "response")

predicted_classes <- ifelse(predicted_probabilities > 0.5, 1, 0)
kable(table(predicted_classes),col.names = c("Value","Frequency"))

```

From this table produced we can see that we get a majority of our predicted values to be 0 and only a few to be 1. This represents the relationship between the bridges that are at risk of collapse that bridges that are not as risk of collapse. We can see that we are getting similar numbers in our prediction to the values contained in the data set. We will now test the accuracy of the model.

```{r}
#| echo: false
#| message: false
#| warning: false
# Actual values
actual_classes <- test_data$collapse_risk
 # Compare predicted vs actual values
comparison <- data.frame(Actual = actual_classes, Predicted = predicted_classes)
 
accuracy <- mean(predicted_classes == actual_classes)
kable(accuracy,col.names = c("Accuracy"))
```

From this table we can see that our model has a 98% accuracy in predicting our defined collapse variable. Which shows us that we can predict with good accuracy if a bridge is at risk of collapse in Omaha,NE.
