---
title: "Methods"
prefer-html: true
---

## Overview of Bridges in Omaha,NE

When thinking about how to conduct the analysis on percent chance a bridge collapses in Omaha, NE, the first step is to find a data set that contains all of the information that is needed. Important information to answer this question would be, how the bridge is structurally, how old it is, and how much traffic the bridge has, for example. The data set that was found is the NTAD National Bridge Inventory, which contains a majority of the information stated above and much more. This data set, described in the data documentation, attempts to figure out the main question at hand what is the percent chance that a bridge will collapse in Omaha, NE. This data set was filtered to contain all bridges in Iowa and Nebraska by state code, it was then filtered again to focus in on the Omaha area. This was done by filter statements in the dyplr package in R, using state codes, latitudes, and longitudes. This subset of the NTAD data was then used for the analysis.

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(psych)
library(knitr)
options(digits = 4)
omaha_bridges <- read.csv("omaha_bridges.csv")
```

```{r}
#| label: fig-omaha-map1
#| fig-cap: "Map of Omaha Bridges"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
library(leaflet)
leaflet(omaha_bridges) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )

```

@fig-omaha-map1 above shows how the data contains all of the bridges in the Omaha, NE area. This is crucial as the task is limited to this area. This map also gives a broader idea of how bridges are distributed across Omaha, NE allowing assessment of the risk for each area. The next step in the analysis is to further understand what variables within the data set are useful. This decision was made by analyzing the code book and understanding which variables give key insight into the structure, age, traffic, and condition of the bridges. These variables will hopefully lead to the ability to draw a conclusion that meet the goal of predicting a bridge collapse.

## Description of Data

```{r}
#| echo: false
#| message: false
#| warning: false

options(digits = 4)
fil_dat <- omaha_bridges %>%
  select(c(1,2,5,14,21,22,25, 28:29,31:32, 68:77,88,103,107,112,116, 122:125))

```

There are 130 variables in the NTAD Bridge Inventory data set, some of variables that will be used are: bridge condition, substructure and superstructure condition, average daily traffic, and year built. These variables give a good picture of what it means to be a stable and healthy bridge and also what it means to be a bridge on the verge of collapse. Next the variable bridge condition was analyzed, which is a variable that describes the overall condition of the bridge, which has rating of "P" for poor, "G" for good, and "F" for fair. Bridge condition rating is determined by, " the lowest rating of National Bridge Inventory (NBI) condition ratings for Item 58 (Deck), Item 59 (Superstructure), Item 60 (Substructure), or Item 62 (Culvert). If the lowest rating is greater than or equal to 7, the bridge is classified as Good; if it is less than or equal to 4, the classification is Poor. Bridges rated 5 or 6 are classified as Fair.", @wesemanRecordingCodingGuide. This allows the analysis of the overall condition of the bridge by looking primarily at one variable to occur.

### Understanding Bridge Condition Variable

```{r}
#| label: fig-omaha-brid
#| fig-cap: "Bridge Condtion Ratings Across Omaha, NE"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
library(ggplot2)

ggplot(fil_dat,aes(x = BRIDGE_CONDITION)) + geom_bar() + labs(title ="Bridge Condition") + xlab("Bridge Rating") + ylab("Number of Bridges")

```

In @fig-omaha-brid we can see the overall breakdown of how the bridge condition variable. When first looking at this graph, it shows that a majority of the bridges in Omaha are in good or fair condition. Next, a closer look was taken into the bridges that have a poor rating.

#### Bridges in Omaha that have a Rating of Poor

```{r}
#| echo: false
#| message: false
#| warning: false
fil_dat <- fil_dat %>%
  mutate(across(everything(), ~ ifelse(. == "N", NA, .)))
```

```{r}
#| label: fig-omaha-map
#| fig-cap: "Map of Omaha Bridges in 'Poor' Condtion"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
bridge_cond <- fil_dat %>%
 filter(
   BRIDGE_CONDITION == "P"
 ) %>%
   select(c(1,4,5,6,8,9:21,29,27,23))


leaflet(bridge_cond) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )
```

@fig-omaha-map above shows the distribution of bridges across Omaha, NE, that have a bridge condition rating of poor. This gives a good understanding of the bridges at most risk of collapse right now. @fig-omaha-map shows that some of the bridges that are at most risk of collapse are in downtown Omaha, which could lead to major economic and logistical issues. The economic impact of a bridge collapse is important for one to consider as it can impact the severity of the incident. Another factor that might play a role on if a bridge collapses would be how much traffic a bridge takes in per day. Another subset of data was created from the bridges that have poor condition, and added that the average daily traffic must be greater than 5,000 cars and that is must have over two lanes. This will give a good picture of major bridges in Omaha, NE that are at risk of collapse that could impact that most people.

#### Bridges in Omaha that have a Rating of Poor and have High Traffic

```{r}
#| label: fig-omaha-map2
#| fig-cap: "Map of Omaha Bridges in 'Poor' Condtion and with High Traffic"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
high_traffic <- bridge_cond %>%
  filter (
    ADT_029 > 5000 & TRAFFIC_LANES_ON_028A > 2
  )
leaflet(high_traffic) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )
```

@fig-omaha-map2 above show the distributed of bridges with a poor condition that undergo high traffic scenarios each day. These bridges could possibly be at higher risk for collapse due to the increase of traffic. The map above shows that none of these bridges are located over water which may mitigate the damage and economic impact of a collapse. The bridges that are at a poor rating and have high traffic are not located in downtown Omaha, which again could limit the economic impact of a collapse.

## Process of Statistical Analysis

Statistical analysis was then conducted, first by analyzing the descriptive statistics of both the bridge condition subset and the high traffic subset. This was then extended by finding the percent of bridges that are at a poor condition and poor condition with high traffic respectively. This gives a basic understanding of what the overall condition currently is. A method of prediction was then needed, which was done by fitting a linear model. The missing values in the data set were a concern which lead to imputation methods being used by way of the "VIM" package in R. This then allowed data to be analyzed using a linear model, creating a method of prediction of when a bridge would collapse in Omaha, NE.
